Applied Measure Theory for Probabilistic Modeling

### 2022-05-17 paper/paper.tex
edits

\section{Base measures and density recursion}
It's very useful for users to be able to \emph{call} \verb|logdensity_rel| for arbitrary measures. But \emph{defining} things directly in these terms would require a definition for every pair of possible measures, which is of course intractable.

@@

The normalization term is present for any normal distribution, even those with a different mean or variance. So if we have a product of normals, for example in a regression problem, the normalization can be shared across them. Taking $n$ iid standard \verb|Normal|s as a simple example, this effectively rewrites the log-pdf
\[
    \sum_{j=1}^n \left(-\frac{1}{2}\log 2π -\frac{x_j^2}{2}\right)
    \quad \text{as} \quad
    -\frac{n}{2}\log 2π + \sum_{j=1}^n \left(-\frac{x_j^2}{2}\right)
\]

Further simplification is clearly possible; doing this efficiently in a composable way is the subject of ongoing work.

@@

\chad{need edits here}

This leaves us with the question of how to compute the three-argument \verb|logdensity(μ, ν, x)| for general $\nu$. We implement this recursively, in an approach we refer to as \emph{density recursion}.

If $\alpha$ is the base measure of $\mu$ and $\beta$ is the base measure of $\nu$, then
\[
\frac{\dd \mu}{\dd \nu} =
\frac{\dd \mu}{\dd \alpha}
\frac{\dd \alpha}{\dd \beta}
\left(\frac{\dd \nu}{\dd \beta}\right)^{-1}\ .
\]

In the implementation, \verb|logdensity(μ, α, x)| and \verb|logdensity(ν, β, x)| can be computed directly using the two-argument \verb|logdensity|. There may be a specific method for \verb|logdensity(α, β, x)|; if not, it will require a recursion step.

In this process, we may reach a measure that serves as its own base measure. Such measures are \emph{primitive}, for example Lebesgue or counting measure. If \verb|logdensity| is called on two primitive measures with no explicit method, an error is thrown.

The above serves as a high-level specification  of the computational process. The low-level implementation will differ in some ways as we continue to tune it for performance.

>>>>

\section{Base measures and log-densities}
It's very useful for users to be able to \emph{call} \verb|logdensity_rel| for arbitrary measures. But \emph{defining} things directly in these terms would require a definition for every pair of possible measures, which is of course intractable.

@@

The normalization term is present for any normal distribution, even those with a different mean or variance. So if we have a product of normals, for example in a regression problem, the normalization can be shared across them. For the simple example of $n$ iid standard \verb|Normal|s, this effectively rewrites the log-pdf
\[
    \sum_{j=1}^n \left(-\frac{1}{2}\log 2π -\frac{x_j^2}{2}\right)
    \quad \text{as} \quad
    -\frac{n}{2}\log 2π + \sum_{j=1}^n \left(-\frac{x_j^2}{2}\right)
\]

Further simplification is clearly possible; doing this in a composable way without sacrificing efficiency is the subject of ongoing work.

@@

Instead of \verb|logdensity_def|, users should typically call \verb|logdensity_rel(μ,ν,x)| (which computes $\frac{\dd μ}{\dd ν}(x)$), or \verb|logdensityof(μ,x)|, which gives the log-density relative to the root measure. \verb|logdensityof| was originally defined in \cite{DensityInterface.jl}.
